Joint estimation of phylogenies and ancestral states is not consistent

### 2017-12-13 appendix.tex
More work on the appendix.

The full Farris tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see Table~\ref{tab:likelihoods} for $\tau = \tau_1$).

@@

Directly substituting in \eqref{eq:site_pattern_likelihood}, the log-likelihood is

@@

@@

The term $p_{\emptyset}$ is bounded as

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
Call
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}. \\
    \end{aligned}
    \label{eq:a_const_prime}
\end{equation}
We also now need to check that
$$
\hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
$$
are both between zero and one.
It will suffice to show
$$
0 \le \tilde{p} - 2p_{13} \le 1
$$
and
$$
0 \le \tilde{p} + 2p_{13} \le 1
$$
for $\tilde{p}$ defined above, which is equivalent to showing $p_{13} \le 1/4$.
Using Table~\ref{tab:sitepatprob} and $t^*$,
$$
p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right).
$$
We see that
$$
1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 2 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - (x^*)^2
$$
and
$$
(y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - 4x^* + (x^*)^2 \le 1 - (x^*)^2,
$$
the last inequality holding if $x^* \le 2$.
Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}$ and $\hat{y}$ are valid.

@@

In the Felsenstein case, there are various optimal internal states depending on branch lengths (Table~\ref{tab:likelihoods}; $\tau = \tau_2$).

@@

@@

%EM Wanna add a short section about numerical optimization? You've seen that it's not entirely trivial.

@@

We initialize our optimization procedure at the true branch parameters $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
We posit that $\hat{w}$ is equal to one for regions of $x^*$ and $y^*$ small; methods that use local optimization may never see candidate values for large $x$ and $y$ if our initial guess is small, so we perform two maximizations, one with $w=1$ fixed and one where $w$ is estimated.
We take the value of $\hat{w}$ with the larger likelihood as our estimate.

>>>>

The full Farris tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see Table~\ref{tab:likelihoods}).

@@

%EM Is eq:site_pattern_likelihood the best ref? It still has counts. It would be nice to have something that's exactly this before plugging in the values from the tables.
Directly substituting calculations from Table~\ref{tab:sitepatprob} and Table~\ref{tab:likelihoods} where $\tau = \tau_1$ in \eqref{eq:site_pattern_likelihood}, the log-likelihood is
%EM This is the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1\}$. You address the other cases later, but at this point it's not clear that you're going to. Can you introduce with this in mind?

@@

1-x^2-y^2-x^2y^2 & = (1+x)(1-x)(1+y)(1-y) \\

@@

The $p_{\emptyset}$ term is bounded as

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
Call
%EM It would seem helful for the below check to express these a_1' etc in terms of tilde{p}.
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}. \\
    \end{aligned}
    \label{eq:a_const_prime}
\end{equation}
%EM This check would make a lot more sense if you put the C^{2}_{\tau_1,\tau_1}(x^*, y^*)  and the other analogous definition first.
%EM It would also be nice to do one of the calculations to get us from a_1 world to a_1' world. Just staring at table S4 doesn't make this obvious.
We also now need to check that
$$
\hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
$$
are both between zero and one.
It will suffice to show
$$
0 \le \tilde{p} - 2p_{13} \le 1
$$
and
$$
0 \le \tilde{p} + 2p_{13} \le 1
$$
for $\tilde{p}$ defined above, which is equivalent to showing $p_{13} \le 1/4$.
Using Table~\ref{tab:sitepatprob} and $t^*$,
$$
p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right).
$$
We see that
\begin{equation}
1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 2 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - (x^*)^2.
\label{eq:lenny}
\end{equation}
Note
$$
(y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - 4x^* + (x^*)^2;
$$
showing that \eqref{eq:lenny} holds if $x^* \le 2$, which is satisfied by definition.
Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}$ and $\hat{y}$ are valid.

@@

In the Felsenstein case, there are quite a few optimal internal states depending on branch lengths (Table~\ref{tab:likelihoods}; $\tau = \tau_2$).

@@

%EM Hm, the following doesn't look right to me.

@@

@@

We initialize our optimization procedure at the true branch parameters $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
We posit that $\hat{w}$ is equal to 1 for when $x^*$ and $y^*$ are small; methods that use local optimization may never see candidate values for large $x$ and $y$ if our initial guess is small, so we perform two maximizations, one with $w=1$ fixed and one where $w$ is estimated.
We take the value of $\hat{w}$ with the larger likelihood as our estimate.
